---
title: "Practical Machine Learning"
author: "Divakar"
date: "Thursday, February 12, 2015"
output: html_document
---

## Summary:
To apply machine learning algorithms to predict human activity recognition . Data from <http://groupware.les.inf.puc-rio.br/har> is used for this purpose
Training data will be divided into two parts 80% for training and 20% for  validating  . Based on outcome of initial models , other models  will be built. Goal is to achieve 90% accuracy . 

### Load data
```{r , echo=TRUE}
library(caret)
library(randomForest)
pml <- read.csv("pml-training.csv" , na.strings=c("","NA"))
dim(pml)

```
### Divide data into two sets 


```{r, echo=TRUE}
trainIndex <- createDataPartition(pml$classe, p = 0.80,list=FALSE)
training <- pml[trainIndex,]
validating <- pml[-trainIndex,]
```
### Explore and clean training data
  
Verify how many fields are NA . Calculate ratio of non-NA to NA  
```{r, echo=TRUE}
ratio_non_na_to_na <- sum(!is.na(training)) / sum(is.na(training)) 
ratio_non_na_to_na

```
There are more NAs than non NA . Keep columns if it contains  less than 30% NA . Drop other columns
```{r, echo=TRUE}
rowCount <- nrow(training)
training <- training[,colSums(is.na(training)) < 0.3 * rowCount]
dim(pml)
```
  
Take a look at and check if some columns can be dropped without causing any issues. By observation of data , X ,user_name ,cvtd_timestamp,new_window columns can be dropped .(output not produced to reduce clutter)

```{r, echo=TRUE}
training <- training[, - c(1,2,5,6)]
dim(training)
```
  
check for near zero variance  
```{r echo=TRUE}
nzv <- nearZeroVar(training[,-c(56)])
nzv
```
There are no nearZero variance columns.  
Identify  highly correlated columns and drop them
```{r , echo=TRUE}
DatCorr <- cor(training[,-c(56)])
highCorr <- findCorrelation(DatCorr, 0.80)
training <- training [,-highCorr]
dim(training)
```
  
With a large set of 43  predictors , random forest gives better result .
  
### Random Forest model with cross validation of 4 fold. Estimate accuracy
```{r , echo=TRUE}
trControl <- trainControl(method = "cv", number = 4)
rf <- train(classe ~ ., method = "rf", trControl = trControl, data=training)
rf
rf$finalModel
plot(varImp(rf, scale = FALSE))

```

  
Based on above data ,Out of sample Error estimated to be 0.1% 
  
### Test model with validating data , create confusionMatrix and verify out of sample accuracy
```{r , echo=TRUE}
predictions <- predict(rf, validating)

confusionMatrix(predictions, validating$classe)


```
  
Accuracy of more than .99 indicate , algorithm worked correctly and out of sample error is within predicted range